---
title: "framgingham project"
author: "MZ"
date: "November 11, 2017"
output: 
  html_document: 
    fig_height: 6
    fig_width: 78
---
# Loading and mutating data
```{r}
library(tidyverse)
#library(dplyr)
library(foreign)
fram <- read.dta(file = "C:/Users/Lem/Dropbox (CBQG)/BST210/project 2/framingham.dta")
#fram <- read.dta(file = "~/BST210/framingham.dta")

fram <- fram %>% mutate(metabolic = ifelse((totchol > 240 & (sysbp > 130 | bpmeds == 1) & bmicat > 2), 1, 0)) %>% filter(!is.na(totchol) & !is.na(sysbp) & !is.na(bpmeds) &!is.na(prevhyp) & !is.na(bmicat) & !is.na(diabetes) & !is.na(prevchd) & !is.na(prevstrk) & !is.na(age) & !is.na(sex) & !is.na(cigpday) & !is.na(cursmoke)) %>% mutate(id = agecat*sex)

dim(fram)
fram %>% filter(diabetes == 1) %>% dim()
fram %>% filter(prevhyp == 1) %>% dim()
fram %>% filter(prevchd == 1) %>% dim()
fram %>% filter(prevstrk == 1) %>% dim()
fram %>% filter(metabolic ==1) %>% dim()

fram <- fram %>% mutate(obese = ifelse((bmicat > 3), 1, 0))
fram <- fram %>% mutate(htc = ifelse((totchol > 240), 1, 0))

#number of pre-existing conditions
fram <- fram %>% mutate(preexisting = diabetes + prevhyp + prevchd + prevstrk, preexisting_m = diabetes + prevhyp + prevchd + prevstrk + metabolic)
fram$preexisting[fram$preexisting==4] <- 2
fram$preexisting[fram$preexisting==3] <- 2
fram$preexisting_m[fram$preexisting_m==4] <- 2
fram$preexisting_m[fram$preexisting_m==3] <- 2

#Table for figure 1
fram %>% group_by(sex) %>% summarize(length(sex)) 

fram %>% group_by(preexisting) %>% summarize(length(preexisting))

fram %>% group_by(preexisting_m) %>% summarize(length(preexisting_m))

fram %>% group_by(preexisting) %>% summarize(mean = mean(cigpday), sd = sd(cigpday))

fram %>% filter(metabolic == 1) %>% summarize(mean = mean(cigpday), sd = sd(cigpday))
```

#Survival Analysis (K-M and log rank test)
```{r}
library(survival)
#Metabolic syndrome
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ metabolic, data = fram)

#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ metabolic, data = fram)

#Hypertension
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ prevhyp, data = fram)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ prevhyp, data = fram)

#Diabetes
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ diabetes, data = fram)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ diabetes, data = fram)

#Obesity
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ obese, data = fram)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ obese, data = fram)

#Angina
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ prevap, data = fram)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ prevap, data = fram)

#CHD
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ prevchd, data = fram)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ prevchd, data = fram)

#smoking
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ cursmoke, data = fram)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ cursmoke, data = fram)

#Pre-existing conditions, excluding metabolic
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ preexisting, data = fram)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ preexisting, data = fram)

#Pre-existing conditions, including metabolic
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ preexisting_m, data = fram)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ preexisting_m, data = fram)

#Part 2: Parametric models
model2 = survreg(survobj ~ sorb+tgh+dur+sex, dist="exponential", data=srt)
```

#ggplot survival curves
```{r}
library(survminer)
#Metabolic syndrome ***
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ metabolic, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE, pval = TRUE, conf.int = TRUE, ggtheme = theme_minimal(), risk.table.fontsize = 3)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ metabolic, data = fram)

#Hypertension
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ prevhyp, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ prevhyp, data = fram)

#Diabetes ***
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ diabetes, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE, pval = TRUE, conf.int = TRUE, ggtheme = theme_minimal(), risk.table.fontsize = 3)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ diabetes, data = fram)

#Obesity
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ obese, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ obese, data = fram)

#Angina
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ prevap, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ prevap, data = fram)

#CHD
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ prevchd, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ prevchd, data = fram)

#smoking
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ cursmoke, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ cursmoke, data = fram)

#Pre-existing conditions, excluding metabolic ****
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ preexisting, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE, pval = TRUE, conf.int = TRUE, ggtheme = theme_minimal(), risk.table.fontsize = 3,  legend.labs = c(0, 1, "2+"), title = "Pre-existing conditions vs Survival")
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ preexisting, data = fram)

#Pre-existing conditions + smoking
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ preexisting + cursmoke, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE, pval = TRUE, conf.int = TRUE, ggtheme = theme_minimal(), risk.table.fontsize = 2, title = "Pre-existing conditions and smoking status vs Survival")
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ preexisting, data = fram)

#Pre-existing conditions, including metabolic
mortality <- Surv(time = fram$timedth, event = fram$death)
model_m <- survfit(mortality ~ preexisting_m, data = fram)
ggsurvplot(model_m, data = fram, risk.table = TRUE)
#KM
plot(model_m,col = c(2,3), xlab = "Days", ylab = "Survival")
#log rank
survdiff(mortality ~ preexisting_m, data = fram)

#Part 2: Parametric models
model2 = survreg(survobj ~ sorb+tgh+dur+sex, dist="exponential", data=srt)
```

#Proportional Hazards Model
```{r}
#Metabolic
model_c = coxph(mortality ~ metabolic + age + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
#with smoking
model_sc = coxph(mortality ~ metabolic + age + sex + cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)

#Hypertension
model_c = coxph(mortality ~ prevhyp + age + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
#with smoking
model_sc = coxph(mortality ~ prevhyp + age + sex + cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)

#Diabetes
model_c = coxph(mortality ~ diabetes + age + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
#with smoking
model_sc = coxph(mortality ~ diabetes + age + sex + cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)

#CHD
model_c = coxph(mortality ~ prevchd + age + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
#with smoking
model_sc = coxph(mortality ~ prevchd + age + sex + cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)


#All *** 
model_c = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + age + sex, data = fram)
plot(survfit(model_c))
summary(model_c)

model_c = coxph(mortality ~ metabolic + preexisting + age + sex, data = fram)
plot(survfit(model_c))
summary(model_c)


#with smoking *** 
model_sc = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + age + sex + cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)

model_sc = coxph(mortality ~ metabolic + preexisting + age + sex + cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)

cox.zph(model_sc)
test_ph <- cox.zph(model_sc)
ggcoxzph(test_ph)

#EM smoking ***
model_sc = coxph(mortality ~ metabolic + metabolic*cigpday + prevhyp + prevhyp*cigpday + diabetes + diabetes*cigpday + prevchd + prevchd*cigpday + prevstrk + prevstrk*cigpday + age + age*cigpday + sex + sex*cigpday + cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + age + sex + sex*cigpday + cigpday, data = fram) # best mod
plot(survfit(model_sc))
summary(model_sc)

model_sc = coxph(mortality ~ metabolic + preexisting + age + sex + cigpday + metabolic*cigpday + preexisting*cigpday + age*cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + preexisting + age + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc)) # best mod
summary(model_sc)

#Confounding of smoking
#smoking could be a potential confounder of number of pre-existing conditions, pre-existing conditions, and metabolic syndrome as it changes the coefficients by > 10%.

```

#GLM (gamma, binomial, poisson, gaussian, inverse.gaussian, quasi, quasibinominal, quasipoisson)
```{r}
summary(glm(death ~ metabolic, data = fram, offset = log(timedth), family = poisson()))
summary(glm(death ~ metabolic, data = fram, family = binomial()))
summary(glm(death ~ prevhyp + diabetes + prevchd + agecat + sex, data = fram, family = binomial()))
summary(glm(death ~ metabolic + prevhyp + diabetes + prevchd, data = fram, offset = log(timedth), family = poisson()))
summary(glm(death ~ agecat + cigpday + sysbp + bmicat + glucose + totchol + sex + heartrte, data = fram, offset = log(timedth), family = poisson()))

```

#Testing proportional hazards assumption (inbetween pre-existing conditions, and smoking vs nonsmoking)
[https://www.r-bloggers.com/cox-model-assumptions/]
```{r}
test_ph <- cox.zph(model_sc)
ggcoxzph(test_ph)
```
In the figure above, the solid line is a smoothing spline fit to the plot, with the dashed lines representing a +/- 2-standard-error band around the fit.

Note that, systematic departures from a horizontal line are indicative of non-proportional hazards, since proportional hazards assumes that estimates  do not vary much over time.

From the graphical inspection, there is no pattern with time. The assumption of proportional hazards appears to be supported for the covariates sex (which is, recall, a two-level factor, accounting for the two bands in the graph), wt.loss and age.

#Outliers
```{r}
ggcoxdiagnostics(model_sc, type = "dfbeta",
                 linear.predictions = FALSE, ggtheme = theme_bw())
```
The above index plots show that comparing the magnitudes of the largest dfbeta values to the regression coefficients suggests that none of the observations is terribly influential individually, even though some of the dfbeta values for age and wt.loss are large compared with the others.

#Stratification with cox model (agecat, pre-existing conditions, and number of pre-existing conditions)
```{r}
library(survival)
model_c = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + agecat + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
#agecat, prevhyp, and prevchd do not satisfy PH

model_c = coxph(mortality ~ metabolic + preexisting + agecat + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
#agecat and preexisting do not satisfy PH

#with smoking *** 
model_sc = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + agecat + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
#agecat, prevhyp, and prevchd do not satisfy PH

model_sc = coxph(mortality ~ metabolic + preexisting + agecat + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
#agecat and preexisting do not satisfy PH

cox.zph(model_sc)

## stratification
model_c = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + I(agecat == 1) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + I(agecat == 2) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + I(agecat == 3) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + I(agecat == 4) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)

model_sc = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + I(agecat == 1) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + I(agecat == 2) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + I(agecat == 3) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + prevhyp + diabetes + prevchd + prevstrk + I(agecat == 4) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)

model_c = coxph(mortality ~ metabolic + I(preexisting == 0) + I(agecat == 1) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 0) + I(agecat == 2) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 0) + I(agecat == 3) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 0) + I(agecat == 4) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 1) + I(agecat == 1) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 1) + I(agecat == 2) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 1) + I(agecat == 3) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 1) + I(agecat == 4) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 2) + I(agecat == 1) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 2) + I(agecat == 2) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 2) + I(agecat == 3) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)
model_c = coxph(mortality ~ metabolic + I(preexisting == 2) + I(agecat == 4) + sex, data = fram)
plot(survfit(model_c))
summary(model_c)

model_sc = coxph(mortality ~ metabolic + I(preexisting == 0) + I(agecat == 1) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 0) + I(agecat == 2) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 0) + I(agecat == 3) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 0) + I(agecat == 4) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 1) + I(agecat == 1) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 1) + I(agecat == 2) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 1) + I(agecat == 3) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 1) + I(agecat == 4) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 2) + I(agecat == 1) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 2) + I(agecat == 2) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 2) + I(agecat == 3) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
model_sc = coxph(mortality ~ metabolic + I(preexisting == 2) + I(agecat == 4) + sex + cigpday + sex*cigpday, data = fram)
plot(survfit(model_sc))
summary(model_sc)
```

#Conditional modeling
```{r}
model_cond <- clogit(death ~ metabolic + preexisting + strata(sex), data=fram)
```

